---
title: "Amelia_workflow"
author: "wisdom & md"
date: "2025-06-16"
output: html_document
---

#### Install Amelia package
```{r eval=FALSE}
install.packages("Amelia")
install.packages("sjlabelled")
```

#### Load Amelia
```{r}
library("Amelia")
library("dplyr")
library("sjlabelled") ## Remove labels from variables
library(tidyr)
library(ggplot2)
```

#### Set seed
```{r}
set.seed(1234)
```

#### Load Data
```{r}  
#Clinical Covariate
cov_data <- readRDS("/restricted/projectnb/cteseq/projects/challenge-project-2024/merged_cte_meta.rds")
```

### Run Amelia

#### Prepare data for Amelia
```{r}
noms=c("npoldd","sleepcb","nppath6","PathMND","Year","Batch","npold","nppath","nplbod","nptdpb","nptdpc","nptdpe","npftdtau","npcort","npprog","npftdt5","npftdt9","npftdtdp","PathAD","PathLBD","PathFTD","CTE","sleepact","suicide","sport","apoe_de","TMEM106B_dom","TMEM106B_invrec","cod","csparCTE","rs1990622","rs3173615","race")#Removed due to not varying in dataset ,,"nppick","npftdt2","npftdt8","npftdt10","PathPrion" 

ords=c("npoldd1","npoldd3","npold1","npold2","npold3","npold4","DementiaHx","ParknismHx","nphemo","npavas","npwmr","nphipscl","npbraak","npneur","npadnc","npdiff","npamy","nparter","CTEStage","Group","apoe","Group_de") #

idvars=c("Core_ID", "SampleName","nppick", "npftdt2", "npftdt8", "npftdt10", "PathPrion", "micdorfront", "micinfpar", "micalc","npoldd2","npoldd4","npoftd")

### Removing variables that do not vary in the dataset
# cov_data_subset <- subset(cov_data, select=-c( nppick, npftdt2, npftdt8, npftdt10, PathPrion, micdorfront, micinfpar, micalc,npoldd2,npoldd4,npoftd))

### Remove labels from variables
cov_data_subset_clean <- remove_all_labels(cov_data)
```

#### Run Amelia on dataset
```{r}
amelia.out1 <- amelia(
                      cov_data_subset_clean, 
                      m=20, 
                      p2s=2, 
                      noms=noms, 
                      ords=ords, 
                      idvars=idvars, 
                      incheck=T,
                      empri=0.3*nrow(cov_data_subset_clean),
                      # parallel = "multicore",
                      # ncpus = 30
                      )

```

### Diagnosis of imputation

#### Histogram of mean imputation (red) vs observed values (blue) of actaully imputed numerical data
```{r}
p <- plot(amelia.out1)
```

#### Overdispersed starting values
```{r}
disperse(amelia.out1, dims = 2, m = 5)
```

### Save imputed datasets
```{r}
filepath <- paste("/restricted/projectnb/cteseq/projects/challenge-project-2024/imputation/Stable_Releases/",Sys.Date(),sep="")
dir.create(filepath)
write.amelia(obj = amelia.out1, file.stem = paste(filepath,"/outdata",sep=""))
```

### Re-imputation for posterior predictive checking

#### Step 1: Duplicate and manipulate the data
```{r}
# Duplicate the data
cov_data_dup <- cov_data_subset_clean

# Set CTE to NA in the duplicated dataset
cov_data_dup$CTE <- NA

# Add a flag to identify original vs duplicate
cov_data_subset_clean$.group <- "original"
cov_data_dup$.group <- "duplicate"

# Combine the datasets
cov_data_reimpute <- bind_rows(cov_data_subset_clean, cov_data_dup)

```

#### Step 2: Re-impute with Amelia
```{r}
library(Amelia)

amelia_reimputed <- amelia(
  cov_data_reimpute,
  m = 2000,
  p2s = 2,
  noms = c(noms),
  ords = ords,
  idvars = c(idvars, ".group"),
  empri = 0.3 * nrow(cov_data_reimpute)
)

```

#### Step 3: Run regression & extract coefficients
```{r}
library(broom)

# Define model formula
model_formula <- as.formula("CTE ~ race + apoe + sport + sleepact")

# Extract betas for original and duplicate imputations
extract_betas <- function(data_list) {
  replicate_coefs <- lapply(data_list$imputations, function(df) {
    df_orig <- df[df$.group == "original", ]
    df_rep  <- df[df$.group == "duplicate", ]
    list(
      orig = coef(lm(model_formula, data = df_orig)),
      rep  = coef(lm(model_formula, data = df_rep))
    )
  })

  # Bind into matrices
  orig_coefs <- do.call(rbind, lapply(replicate_coefs, `[[`, "orig"))
  rep_coefs  <- do.call(rbind, lapply(replicate_coefs, `[[`, "rep"))
  
  list(original = orig_coefs, replicate = rep_coefs)
}

coef_data <- extract_betas(amelia_reimputed)

```

#### Step 4: Calculate PPP
```{r}
# Function to calculate PPP
compute_ppp <- function(orig_mat, rep_mat) {
  apply(abs(rep_mat) >= abs(orig_mat), 2, mean)
}

ppp_vals <- compute_ppp(coef_data$original, coef_data$replicate)
print(ppp_vals)

```

#### Step 5: Plot distributions
```{r}
# Stack for plotting
coefs_df <- as.data.frame(coef_data$replicate)
coefs_df_long <- pivot_longer(coefs_df, everything(), names_to = "term", values_to = "estimate")

# Original means
orig_means <- colMeans(coef_data$original)
orig_df <- data.frame(term = names(orig_means), estimate = orig_means)

ggplot(coefs_df_long, aes(x = estimate)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  geom_vline(data = orig_df, aes(xintercept = estimate), color = "red", linetype = "dashed") +
  facet_wrap(~term, scales = "free") +
  labs(title = "Posterior Predictive Check: Distribution of Coefficients",
       subtitle = "Red dashed line: Original estimate") +
  theme_minimal()

```

#### Optional Scatter plot
```{r}
# Prepare data frame for scatterplot
scatter_df <- data.frame(
  term = colnames(coef_data$original),
  orig_mean = colMeans(coef_data$original),
  rep_mean  = colMeans(coef_data$replicate)
)

# Make scatterplot
ggplot(scatter_df, aes(x = orig_mean, y = rep_mean, label = term)) +
  geom_point(size = 3) +
  geom_text(nudge_y = 0.05, check_overlap = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") + 
  theme_minimal() +
  labs(
    x = "Mean Coefficient (Original Completed Data)",
    y = "Mean Coefficient (Re-Imputed Data)",
    title = "PPC Scatterplot: Original vs Re-Imputed Coefficients",
    subtitle = "Dashed line = perfect agreement"
  )


```

